Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-23 manuscript/manuscript.Rmd
specify lno doi

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network \citep{covino2017}. For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance \citep{gomi2002}. Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

We asked two questions in this study: 1) Which measures of freshwater connectivity influence lake phosphorus retention? 2) What spatial extent of connectivity most strongly influences P retention? To answer these questions, we fit statistical P retention models in a Bayesian hierarchical framework following Cheng et al. (2010) where 2 separate values of the k processing parameter were estimated for lakes with either high or low values of each connectivity metric. Using this approach, higher k values for a specific lake connectivity class indicates more extensive in-lake processing and higher P retention. We applied this model to a dataset of 129 lakes across a wide range of hydrologic, geologic, and climatic settings. We fit separate models using each combination of connectivity metric and spatial extent in an effort to determine whether P retention is more strongly controlled at the lake subwatershed extent or the lake watershed extent.

@@

We used data on P retention, maximum depth, and water residence time from 129 lakes in the National Eutrophication Survey (USEPA 1975, Stachelek et al. 2018). Mean annual P loading, P discharge, and P retention values in the National Eutrophication Survey (NES) dataset were calculated based on monthly sampling for P in tributary and outlet discharge points as well as any municipal waste discharges from 1972 to 1975. Here, P retention is a unitless value representing the fraction of incoming P loads. Sampling frequency for water discharge and residence time varied among lakes but details of these variations were not provided in the source dataset (USEPA 1975, Stachelek et al. 2018). Estimates of water discharge and residence time in the NES dataset represent normalized mean flow estimates expected to occur during a period of average precipitation and hydrology (USEPA 1975). Water residence time for our study lakes ranged from 1 week to 17 years with an interquartile range of 3 months to 1.8 years while P retention ranged from 0.06 to 0.99 with an interquartile range of 0.24 to 0.59 (Table 1).

@@

We supplemented the NES dataset with boundaries for lake subwatersheds, as well as estimates of stream density, upstream lake area, upstream lake connection(s), baseflow (an index of groundwater inputs), land-use cover, and other water quality measurements from the LAGOS-NE dataset (Table 1; Soranno et al. 2017). Our study lakes encompassed a wide range of land-use cover types and nutrient levels (Table 1). Although, lake subwatersheds were variable with respect to agricultural land use cover, we did not observe a strong relationship with lake P retention (Figure A2). On average, the water quality (total phosphorus, chlorophyll concentration, and Secchi depth) of the lakes in our study are similar to other US lakes as measured by the stratified random sampling design of the National Lakes Assessment (NLA) lake population (USEPA 2016). However, our lakes are substantially larger and deeper than most NLA lakes (Figure A3).

We restricted the lakes in the study to those located within the footprint of LAGOS-NE which includes lakes located in 17 northeastern and midwestern US states (Soranno et al. 2017). We excluded lakes from our analysis if they had a surface area of greater than 1000 km2 or a surface area of less than 0.1 km2. We also excluded lakes if they had a maximum depth of greater than 70 m, lacked upstream surface water connections, or had one of the North American Great Lakes in its upstream watershed. A total of 129 out of 236 NES lakes met each of these selection criteria.

@@

In addition to data from the NES and LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity (Rodriguez-Iturbe and Rinaldo 2001). In addition to stream-based metrics, we calculated lake-stream-based metrics that we expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” (Fergus et al. 2017). In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal to 10 ha (0.1 $km^{2}$). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input \citep{usgsNationalHydrographyDataset2018}. Average link length was calculated as the total stream length in a given watershed divided by the number of stream reaches after dissolving (removing) any network points that do not occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes) expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams \citep{labarberaFractalDimensionStream1989}. Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively \citep{R-streamnet, R-nhdR}. The algorithms in the `streamnet` package use the `sf` R package \citep{pebesma2018} as well as the `v.net` and `v.stream.order` modules \citep{jasiewiczNewGRASSGIS2011} included in GRASS GIS \citep{GRASS_GIS_software}. All processed connectivity data and code are available at [Zenodo DOI].

>>>>

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network \citep{covino2017}. For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance \citep{gomi2002}. Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes \citep{cardilleCarbonWaterCycling2007}. In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

We asked two questions in this study: 1) Which measures of freshwater connectivity influence lake phosphorus retention? 2) What spatial extent of connectivity most strongly influences P retention? To answer these questions, we fit statistical P retention models in a Bayesian hierarchical framework following \citet{cheng2010} where 2 separate values of the $k$ processing parameter were estimated for lakes with either high or low values of each connectivity metric. Using this approach, higher k values for a specific lake connectivity class indicates more extensive in-lake processing and higher P retention. We applied this model to a dataset of 129 lakes across a wide range of hydrologic, geologic, and climatic settings. We fit separate models using each combination of connectivity metric and spatial extent in an effort to determine whether P retention is more strongly controlled at the lake subwatershed extent or the lake watershed extent.

@@

We used data on P retention, maximum depth, and water residence time from 129 lakes in the National Eutrophication Survey \citep{usepa1975, stachelek_national_2018}. Mean annual P loading, P discharge, and P retention values in the National Eutrophication Survey (NES) dataset were calculated based on monthly sampling for P in tributary and outlet discharge points as well as any municipal waste discharges from 1972 to 1975. Here, P retention is a unitless value representing the fraction of incoming P loads. Sampling frequency for water discharge and residence time varied among lakes but details of these variations were not provided in the source dataset \citep{usepa1975, stachelek_national_2018}. Estimates of water discharge and residence time in the NES dataset represent normalized mean flow estimates expected to occur during a period of average precipitation and hydrology \citep{usepa1975}. Water residence time for our study lakes ranged from 1 week to 17 years with an interquartile range of 3 months to 1.8 years while P retention ranged from 0.06 to 0.99 with an interquartile range of 0.24 to 0.59 (Table 1).

@@

We supplemented the NES dataset with boundaries for lake subwatersheds, as well as estimates of stream density, upstream lake area, upstream lake connection(s), baseflow (an index of groundwater inputs), land-use cover, and other water quality measurements from the LAGOS-NE dataset \citep[Table 1; ]{sorannoLAGOSNEMultiscaledGeospatial2017}. Our study lakes encompassed a wide range of land-use cover types and nutrient levels (Table 1). Although, lake subwatersheds were variable with respect to agricultural land use cover, we did not observe a strong relationship with lake P retention (Figure A2). On average, the water quality (total phosphorus, chlorophyll concentration, and Secchi depth) of the lakes in our study are similar to other US lakes as measured by the stratified random sampling design of the National Lakes Assessment (NLA) lake population \citep{usepa_nla2012}. However, our lakes are substantially larger and deeper than most NLA lakes (Figure A3).

We restricted the lakes in the study to those located within the footprint of LAGOS-NE which includes lakes located in 17 northeastern and midwestern US states \citep{sorannoLAGOSNEMultiscaledGeospatial2017}. We excluded lakes from our analysis if they had a surface area of greater than 1000 $km^{2}$ or a surface area of less than 0.1 $km^{2}$. We also excluded lakes if they had a maximum depth of greater than 70 m, lacked upstream surface water connections, or had one of the North American Great Lakes in its upstream watershed. A total of 129 out of 236 NES lakes met each of these selection criteria.

@@

In addition to data from the NES and LAGOS-NE datasets, we calculated several connectivity metrics that we expected would be related to lake P retention (Figure 2). Some of these metrics were stream-based with the goal of capturing aspects of the configuration of each lakes’ upstream surface water network (Figure 2C-D). In particular, we chose metrics that would quantitatively approximate network complexity under the assumption that highly complex networks are also low connectivity networks. This assumption is supported by the findings of stream network simulations where increased network complexity leads to increased network resistance and ultimately decreases in network connectivity \citep{rodriguez2001fractal}. In addition to stream-based metrics, we calculated lake-stream-based metrics that we expected would reflect the likelihood of P trapping in upstream lakes prior to arriving at a given focal lake via tributary flow (Figure 2A, 2B). A simple metric that captures this likelihood is the presence (or absence) of an upstream lake (greater than 4 ha) which we define as “lake connection” \citep{fergusFreshwaterLandscapeLake2017}. In addition to lake connection, we calculated related metrics such as total upstream lake area and the network distance to the closest upstream lake.
To examine the importance of spatial extent relative to our connectivity metrics, we calculated each metric at multiple extents (Figure 3). First, we calculated connectivity metrics at the scale of individual lake subwatersheds. We defined a lake subwatershed as the area draining into a particular lake exclusive of any upstream areas that drain into a lake greater than or equal to 10 ha (0.1 $km^{2}$). Next, we calculated connectivity metrics at the scale of entire upstream lake networks (lake watershed extent). We defined a lake watershed as the area draining into any part of the upstream network irrespective of the presence or absence of upstream lakes (Figure 3).
All connectivity metrics were calculated using the high-resolution National Hydrography Dataset (NHD) as a primary input \citep{usgsNationalHydrographyDataset2018}. Average link length was calculated as the total stream length in a given watershed divided by the number of stream reaches after dissolving (removing) any network points that do not occur at a stream junction. Stream density was calculated as the length of all streams in the watershed (minus artificial lines through lakes) expressed in units of meters per hectare. Upstream lake area was calculated as the sum of the lake area in the upstream watershed expressed in square meters. Stream order ratio was defined as the number of headwater (first-order) streams in the upstream watershed of the focal lake divided by the total number of higher order (> 1) streams \citep{labarberaFractalDimensionStream1989}. Closest distance to an upstream lake was defined as the shortest path-distance (rather than the straight-line distance) to a lake upstream from the focal lake. We calculated all connectivity metrics and lake watershed extents using the `streamnet` and `nhdR` R packages respectively \citep{R-streamnet, R-nhdR}. The algorithms in the `streamnet` package use the `sf` R package \citep{pebesma2018} as well as the `v.net` and `v.stream.order` modules \citep{jasiewiczNewGRASSGIS2011} included in GRASS GIS \citep{GRASS_GIS_software}. All processed connectivity data and code are available at [Zenodo DOI].
